---
title: "Metabolite Set Enrichment Analysis (MSEA) for metabolomics community"
author:
- name: Atsushi Fukushima
  affiliation: RIKEN Center for Sustainable Resource Science
  email: afukushima@gmail.com
- name: Kozo Nishida
  affiliation: RIKEN Quantitative Biology Center
  email: knishida@riken.jp
date: "`r Sys.Date()`"  
output:
  BiocStyle::html_document2:
    toc_float: true
  BiocStyle::pdf_document2: default
package: MSEAp
abstract: |
  Instructions for the MSEAp package.
vignette: |
  %\VignetteIndexEntry{MSEAp instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
One of the ultimate goals in systems biology is to elucidate the genotype-phenotype relationship in cellular systems [1]. Integrated 'omics' approaches have received particular attention. Most omics analyses monitor the levels of target molecules like transcripts and metabolites. In this context, enrichment analysis approaches can be combined with pathway analysis to evaluate whether a particular molecular group is significantly over-represented, for example Gene Set Enrichment Analysis (GSEA) [2-3]. In meteabolomics, one of promising approaches is [Metabolite Set Enrichment Analysis (MSEA)](https://academic.oup.com/nar/article-lookup/doi/10.1093/nar/gkq329) [4]. 
MSEAp is an R package for performing MSEA using both plant and animal metabolic pathways and metabolite-sets. The package can import pre-defined pathways like KEGG and SMPDB and user-defined metabolite sets in [gmt format](http://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats). Their metabolite sets are integrated and used for MSEA calculation. Given a metabolite-set or pathway annotation of metabolites, significance of the over-representation of the metabolites within the pathways is determined by a Fisher's exact test. The barplot visualization is based on previously established, a web server MSEA in [MetaboAnalyst](http://www.metaboanalyst.ca/) [5]. The package also contains pre-defined and extracted AraCyc that are publicly available and our curated metabolite-sets from MeKO [6], Arabidopsis flavonoid pathways (Tokimatsu, Nishida et al., in preparation), and literatures in plant metabolomics (AtMetExpress, in preparation).


# Installation
```R
## under construction
install.packages("devtools")

source("http://bioconductor.org/biocLite.R")
biocLite(c("KEGGREST", "KEGGgraph"))

library(devtools)
install_github("afukushima/MSEAp")
```

# Getting started

The following is an example of MSEA.
`mset_SMPDB_format_KEGG` is a metabolite-set that has the metabolites for each pathway from the [SMPDB](http://smpdb.ca) [7] database with the metabolite ID from the [KEGG](http://www.genome.jp/kegg/) [8] database.

`kusano` has KEGG metabolite IDs for MESA.

`res` is the MSEA result and the `plot` function visualizes the top 20 over-represented pathways.

```{r}
library(MSEAp)
data(mset_SMPDB_format_KEGG)
data(kusano)
res <- msea(mset_SMPDB_format_KEGG, kusano)
plot(res)
##
head(kusano)
mset_SMPDB_format_KEGG[[1]]
```

# Appetizer

## Importing user's defined metabolite-sets
You can import user's defined metabolite-sets in [*.gmt format](http://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats) using read.gmt() function:
```{r}
file <- system.file("extdata", "sample_SMPDB.gmt", package = "MSEAp")
smp <- read.gmt(file)
```

## Importing KGML file 
The function kgml.import() can extract compound set from [KGML](http://www.kegg.jp/kegg/xml/) file (*.xml) to list.
```{r, warning = FALSE}
file <- system.file("extdata/map00260.xml", package="KEGGgraph")## KGML downloaded from KEGG ftp
kgml.import(file)
```

## Importing KEGG pathways
The function setup.KEGG.gmt.R() generates metabolite-sets using the [KEGG REST API](http://www.kegg.jp/kegg/docs/keggapi.html).
```{r}
res <- setup.KEGG.gmt(listformat = TRUE)
res[[1]]
```

## Conversion of KEGG's tar.gz into metabolite-set
The function setupPathwayClass() performs convertion of KEGG's tar.gz files (e.g. hsa.tar.gz) to metabolite-set list.
```{r}
## filepath <- "./hsa.tar.gz" ## *.tar.gz downloaded from KEGG
## res <- setupPathwayClass(filepath) ## convert to list
```

## Metabolite-sets
You can see the currently available datasets:
```{r}
supported.msets()
```

Below is the selected table of the number of our metabolite-sets.

| Data source   | KEGG | HMDB | KNApSAcK | 
| :--- | ---: | ---: | ---: | ------------- |
| SMPDB         | 717  | 717  | NA |
| AraCyc        | 268  | 350  | NA |
| PlantCyc      | 606  | 714  | NA |
| AtMetExpress (Stress)| 9    | NA | NA |
| AtMetExpress (2nd Metabolism) | NA | NA | 4 |


## For plant scientists
You can perform MSEA using AraCyc pathways and our curated metabolite sets (flavonoid pathways and some litterature).
```{r}
data(mset_AraCyc_format_KEGG)
data(mset_AtMetExpress_Stress_format_KEGG)
data(mset_AtMetExpress_Flavonoids_format_KNApSAcK)
mset <- c(
  mset_AraCyc_format_KEGG,
  mset_AtMetExpress_Stress_format_KEGG,
  mset_AtMetExpress_Flavonoids_format_KNApSAcK
  )
length(mset) ## 281 metabolite-sets
data(fukushima17_INC)
head(fukushima17_INC)
res <- msea(mset, fukushima17_INC$kegg_knap)
head(res)
plot(res)
```


# Visualization of MSEA results
To help interpreting MSEA result, we also implemented `barplot`, `dotplot` and `netplot` for visualization.

`barplot` plots pvalue as color and the number of metabolite queries contained in the metabolite group as the height of bar.

`dotplot` plots dots on the expected value of msea in each group, with pvalue as color (like barplot) and the number of metabolite queries contained in the metabolite group as the size of dot.

`netplot` receives the edge list written with `write.network` beforehand and plots the pvalue in the MSEA result as the node color on the network of the metabolite group.

The option shared.metabolite in `write.network` decides how to connect the metabolite group with the edge.

In the example below, if there are more than 20 metabolites common among groups, we connect these groups with edge.

```{r}
# data(mset_SMPDB_Metabolic_format_HMDB)
# data(msea.example)
# res <- msea(mset_SMPDB_Metabolic_format_HMDB, msea.example)
# 
# dotplot(res)  ## You may have to zoom this plot in RStudio and other R working environments
# barplot(res)  ## You can see the same plot by plot()
# 
# netplot(res, mset_SMPDB_Metabolic_format_HMDB)
res <- msea(mset_SMPDB_format_KEGG, kusano)
dotplot(res)  ## You may have to zoom this plot in RStudio and other R working environments
barplot(res)  ## You can see the same plot by plot()

netplot(res, mset_SMPDB_format_KEGG, shared.metabolite = 20)

```


# Acknowledgments

We thank Dr. Toshiaki Tokimatsu in NIG for manual curation of the flavonoid biosynthetic pathway in _Arabidopsis thaliana_.


# References
1. Fukushima A, Kanaya S, Nishida K: Integrated network analysis and effective tools in plant systems biology. [Front Plant Sci. 2014 Nov 4;5:598. doi: 10.3389/fpls.2014.00598](https://dx.doi.org/10.3389/fpls.2014.00598).
2. Mootha VK, Lindgren CM, Eriksson KF, Subramanian A, Sihag S, Lehar J, Puigserver P, Carlsson E, Ridderstråle M, Laurila E, Houstis N, Daly MJ, Patterson N, Mesirov JP, Golub TR, Tamayo P, Spiegelman B, Lander ES, Hirschhorn JN, Altshuler D, Groop LC: PGC-1alpha-responsive genes involved in oxidative phosphorylation are coordinately downregulated in human diabetes. [Nat Genet. 2003 Jul;34(3):267-273](http://www.nature.com/ng/journal/v34/n3/abs/ng1180.html).
3. Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, Paulovich A, Pomeroy SL, Golub TR, Lander ES, Mesirov JP: Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. [Proc Natl Acad Sci U S A. 2005 Oct 25;102(43):15545-15550](http://www.pnas.org/content/102/43/15545.abstract).
4. Xia J and Wishart DS: MSEA: a web-based tool to identify biologically meaningful patterns in quantitative metabolomic data. [Nucleic Acids Res. 2010 Jul;38(Web Server issue):W71-77. doi: 10.1093/nar/gkq329](https://academic.oup.com/nar/article-lookup/doi/10.1093/nar/gkq329).
5. Xia J, Sinelnikov IV, Han B, Wishart DS: MetaboAnalyst 3.0--making metabolomics more meaningful. [Nucleic Acids Res. 2015 Jul 1;43(W1):W251-257. doi: 10.1093/nar/gkv380](https://academic.oup.com/nar/article-lookup/doi/10.1093/nar/gkv380).
6. Fukushima A, Kusano M, Mejia RF, Iwasa M, Kobayashi M, Hayashi N, Watanabe-Takahashi A, Narisawa T, Tohge T, Hur M, Wurtele ES, Nikolau BJ, Saito K: Metabolomic Characterization of Knockout Mutants in Arabidopsis: Development of a Metabolite Profiling Database for Knockout Mutants in Arabidopsis. [Plant Physiol. 2014 Jul;165(3):948-961](http://www.plantphysiol.org/content/165/3/948.long).
7. Jewison T, Su Y, Disfany FM, Liang Y, Knox C, Maciejewski A, Poelzer J, Huynh J, Zhou Y, Arndt D, Djoumbou Y: SMPDB 2.0: big improvements to the Small Molecule Pathway Database. [Nucleic acids research. 2013 Nov 6;42(D1):D478-84](https://academic.oup.com/nar/article/42/D1/D478/1047198).
8. Kanehisa M, Furumichi M, Tanabe M, Sato Y, Morishima K: KEGG: new perspectives on genomes, pathways, diseases and drugs. [Nucleic acids research. 2017 Jan 4;45(D1):D353-61](https://academic.oup.com/nar/article/45/D1/D353/2605697).


# Session info {.unnumbered}

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
